Markov Model

### 2025-12-03 analysis/discrete-markov.qmd
Update :fire:

```{r}
#| label: covariates
#|
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("apathy_", seq(2016, 2022, by = 2)),
  paste0("mean_anxiety_", seq(16, 22, by = 2)),
  "Total_p")
```

@@

fit_a <- nnet::multinom(
  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
  family = multinomial,
  data = data_stack, trace = FALSE)

fit_b <- nnet::multinom(
  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
  family = multinomial,
  data = data_stack |> mutate(status = relevel(status, ref = 2)),
  trace = FALSE)
```

@@

interaction_results <- rbind(tidy_output(fit_a), tidy_output(fit_b)) |>
  mutate(
    # Mutating to factors
    transition = factor(
      transition,
      levels = c("NC - MCI", "MCI - NC",
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term,
      levels = c("Being female", "Age",
                 "High school degree vs. No education",
                 "Further education vs. No education",
                 "Apathy Scores", "Anxiety Scores",
                 "Procrastination (2020)",
                 "Age x Procrastination", "Previous state: MCI",
                 "Previous state: Dementia")),
    )

@@

# Design matrix for faceting
design <- "
  AABBCC
  DDEEFF
  ##GG##
"

@@

# Making a new predicted dataset
pred_data <- expand.grid(
  Gender = factor(0),
  Age = seq(60, 97, length = 200),
  Education_tri = factor(0),
  Apathy = mean(data_stack$Apathy, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

@@

# Create predictions with age info
age_cohots <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Apathy = mean(data_stack$Apathy, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200)) |>
  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") &
         status %in%  c("Normal Cognition", "MCI")) |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age),
    TRUE ~ "Other"
  ))

@@

# Plotting ---------------------------------------------------------------------
ggplot(data = age_cohots, aes(x = Total_p, y = prob,
                              group = interaction(Age, status_prev, status),
                              colour = label)) +
  geom_line(aes(linewidth = label)) +
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = paste0("Age ", Age), colour = as.character(Age)),
    direction = "y", nudge_x = 4, hjust = 0,
    segment.size = 0.5, segment.colour = "grey50",
    size = 3, show.legend = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
  scale_colour_manual(values = c(
    "70"    = "#E69F00",
    "80"    = "#56B4E9",
    "90"    = "#009E73",
    "Other" = "grey80")) +
  scale_linewidth_manual(values = c(
    "70"    = 1.5,
    "80"    = 1.5,
    "90"    = 1.5,
    "Other" = 0.5)) +
  labs(
    # title = "Predicted transition probabilities for different age cohorts",
    x = "Procrastination", y = "Probability", colour = NULL) +
  facet_grid(status ~ status_prev, labeller = labeller(
    status = c("Normal Cognition" = "Normative Cognitive Function"),
    status_prev = c("Normal Cognition" = "Normative Cognitive Function")
  )) +
  theme_clean() +
  theme(
    panel.grid = element_blank()
    legend.title = element_text(size = 10, face = "bold"),
    legend.position = "none")
```

>>>>

```{r}
#| label: covariates
#|
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("apathy_", seq(2016, 2022, by = 2)),
  paste0("depression_", seq(2016, 2022, by = 2)),
  paste0("mean_anxiety_", seq(16, 22, by = 2)),
  "Total_p")
```

@@

fit_a <- nnet::multinom(
  status ~ Gender + Education_tri + Apathy + Depression + (Total_p * Age) + status_prev,
  family = multinomial,
  data = data_stack, trace = FALSE)

fit_b <- nnet::multinom(
  status ~ Gender + Education_tri + Apathy + Depression + (Total_p * Age) + status_prev,
  family = multinomial,
  data = data_stack |> mutate(status = relevel(status, ref = 2)),
  trace = FALSE)
```

@@

interaction_results <- rbind(tidy_output(fit_a), tidy_output(fit_b)) |>
  mutate(
    # Mutating to factors
    transition = factor(
      transition,
      levels = c("NC - MCI", "MCI - NC",
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term,
      levels = c("Being female", "Age",
                 "High school degree vs. No education",
                 "Further education vs. No education",
                 "Apathy", "Anxiety", "Depression",
                 "Procrastination (2020)",
                 "Age x Procrastination", "Previous state: MCI",
                 "Previous state: Dementia")),
    )

@@

# Design matrix for faceting
design <- "
  AABBCC
  DDEEFF
  #GGHH#
"

@@

# Making a new predicted dataset
pred_data <- expand.grid(
  Gender = factor(0),
  Age = seq(60, 97, length = 200),
  Education_tri = factor(0),
  Apathy = mean(data_stack$Apathy, na.rm = TRUE),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

@@

# Create predictions with age info
age_cohots <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Apathy = mean(data_stack$Apathy, na.rm = TRUE),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200)) |>
  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") &
         status %in%  c("Normal Cognition", "MCI")) |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age),
    TRUE ~ "Other"
  ))

@@

# Plotting ---------------------------------------------------------------------
ggplot(data = age_cohots, aes(x = Total_p, y = prob,
                              group = interaction(Age, status_prev, status),
                              colour = label)) +
  geom_line(aes(linewidth = label)) +
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = paste0("Age ", Age), colour = as.character(Age)),
    direction = "y", nudge_x = 4, hjust = 0,
    segment.size = 0.5, segment.colour = "grey50",
    size = 3, show.legend = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
  scale_colour_manual(values = c(
    "70"    = "#E69F00",
    "80"    = "#56B4E9",
    "90"    = "#009E73",
    "Other" = "grey80")) +
  scale_linewidth_manual(values = c(
    "70"    = 1.5,
    "80"    = 1.5,
    "90"    = 1.5,
    "Other" = 0.5)) +
  labs(
    # title = "Predicted transition probabilities for different age cohorts",
    x = "Procrastination", y = "Probability", colour = NULL) +
  facet_grid(status ~ status_prev, labeller = labeller(
    status = c("Normal Cognition" = "Normative Cognitive Function"),
    status_prev = c("Normal Cognition" = "Normative Cognitive Function")
  )) +
  theme_clean() +
  theme(
    panel.grid = element_blank(),
    legend.title = element_text(size = 10, face = "bold"),
    legend.position = "none") -> new_fig
```
